---
title: "Extreme Value Analysis and Portfolio Optimization for US Equities"
subtitle: "Stable, GEV, GPD, and copula models applied to a sector-diversified basket"
author: "Daniel Bereczki"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  cache     = TRUE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)
set.seed(20240518)

## Where README figures are written (created by the final chunk).
FIG_DIR <- "figures"
dir.create(FIG_DIR, showWarnings = FALSE)
```

# Introduction {-}

This report applies univariate and bivariate extreme-value techniques,
copula modelling, portfolio optimization, and the PSY bubble test to a
sector-diversified basket of ten US large-cap equities. The methodology
mirrors a classic *price fluctuations* course workflow but adapted to
freely downloadable Yahoo Finance data.

The methods used are:

* stable distribution fitting via `stableFit` (quantile and MLE estimators),
* GEV fitting on block maxima via `blockMaxima` / `gev.fit` / `gevFit`,
* GPD fitting via the Peaks-Over-Threshold method (`gpd.fit`, `gpdFit`)
  with the built-in `evd::mrlplot` and `evd::tcplot` for threshold
  selection,
* Michael's stabilized PP-plot,
* profile-likelihood-based confidence intervals via `gev.prof` and
  `gpd.prof`,
* a bivariate extreme-value distribution with Fréchet margins via `fbvevd`,
* copula fitting and goodness-of-fit (`fitCopula`, `gofCopula`, Rosenblatt
  transform `rtrafo`),
* portfolio optimization with `tseries::portfolio.optim` and
  `fPortfolio::min*Portfolio`,
* the built-in `ks.test`, `boot::boot`, and `boot::boot.ci` for
  goodness-of-fit and bootstrap intervals.

Two short helper functions remain in the report (a chi-squared test on
stable-quantile bins and a chi-squared test for GEV) because no built-in
implementation applies the correct degrees-of-freedom correction for
estimated parameters.

# Packages and data

Install any missing packages using the optional block below.

**System dependency.** The `copula` package requires the `gsl` R package,
which in turn needs the GNU Scientific Library at the OS level (the error
*"there is no package called 'gsl'"* signals this is missing). Install
first:

* Linux (Debian/Ubuntu): `sudo apt-get install libgsl-dev`
* macOS (Homebrew): `brew install gsl`
* Windows: the CRAN binary of `gsl` bundles the library, so
  `install.packages("gsl")` normally suffices.

After the system library is installed, run `install.packages(c("gsl",
"copula"))` so that `copula` links against it.

**Note on `psymonitor`.** The `psymonitor` package was archived on CRAN.
Install from the CRAN archive (requires Rtools on Windows):

```r
install.packages(
  "https://cran.r-project.org/src/contrib/Archive/psymonitor/psymonitor_0.0.2.tar.gz",
  repos = NULL, type = "source"
)
```

```{r install-packages, eval=FALSE}
need <- c(
  "fBasics", "stabledist", "fExtremes", "evir", "ismev", "evd",
  "POT", "gsl", "copula", "boot", "tseries", "fPortfolio", "Rglpk",
  "timeSeries", "fitdistrplus", "goftest", "ADGofTest",
  "psymonitor", "knitr"
)
to_install <- setdiff(need, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install)
```

```{r packages}
library(fBasics)        # stableFit
library(stabledist)     # dstable / pstable / qstable / rstable
library(fExtremes)      # blockMaxima, gevFit, gpdFit, dgev/qgev, dgpd/qgpd
library(evir)           # alternative gev parameterization
library(ismev)          # gev.fit, gev.diag, gev.prof, gpd.fit, gpd.diag, gpd.prof
library(evd)            # rmvevd / rbvevd / fbvevd, abvnonpar, qfrechet/pfrechet
library(POT)            # POT-style utilities
library(copula)         # fitCopula, gofCopula, rtrafo, pobs
library(boot)           # bootstrap, bca-type confidence intervals
library(tseries)        # portfolio.optim
library(fPortfolio)     # minvariancePortfolio / minriskPortfolio
library(timeSeries)     # timeSeries class
library(psymonitor)     # PSY bubble test
library(knitr)
```

## Loading the Yahoo Finance data

Prices come from `data/stock_prices.csv`, produced by `python
data_download.py`. That script downloads adjusted daily closes with
`yfinance`, so splits and dividends are already handled.

```{r load-data}
if (!file.exists("data/stock_prices.csv")) {
  stop("data/stock_prices.csv not found -- run `python data_download.py` first.")
}

prices_raw <- read.csv("data/stock_prices.csv", stringsAsFactors = FALSE)
dates <- as.Date(prices_raw$Date)
prices <- prices_raw[, setdiff(names(prices_raw), "Date")]
prices <- na.omit(prices)
prices <- as.data.frame(prices)
T_full <- nrow(prices)
cat("Trading days with prices for all tickers:", T_full, "\n")
cat("Tickers:", paste(colnames(prices), collapse = ", "), "\n")

## daily log-returns ---------------------------------------------------------
returns <- as.data.frame(apply(prices, 2, function(x) diff(log(x))))
losses  <- -returns                  # losses = negative returns
n        <- nrow(returns)
n_stocks <- ncol(returns)
stocks   <- colnames(returns)
cat("Log-return observations per stock:", n, "\n")
```

## Descriptive statistics

```{r descriptive}
desc <- data.frame(
  Mean       = sapply(returns, mean),
  SD         = sapply(returns, sd),
  Skewness   = sapply(returns, function(x) mean((x-mean(x))^3)/sd(x)^3),
  Kurtosis   = sapply(returns, function(x) mean((x-mean(x))^4)/sd(x)^4),
  Min        = sapply(returns, min),
  Max        = sapply(returns, max),
  Emp99Loss  = sapply(returns, function(x) -quantile(x, 0.01))
)
kable(round(desc, 5), caption = "Descriptive statistics of daily log-returns.")
```

```{r return-plots, fig.width=10, fig.height=6}
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:10) {
  plot(returns[, j], type = "l", col = "steelblue",
       main = stocks[j], xlab = "", ylab = "", cex.main = 0.9)
  abline(h = 0, col = "grey")
}
par(op)
```

Every series shows the well-known stylized facts of equity returns:
approximately zero mean, heavy tails, and volatility clustering.
Empirical excess kurtosis is well above 0 for every stock, so
heavy-tailed models are needed.

# Task 1 -- Univariate analysis

## Stable distribution fit (`stableFit`)

The four-parameter $\alpha$-stable distribution is fitted with
**McCulloch's quantile estimator** (`type="q"`) and then refined by **MLE**
(`type="mle"`).

```{r stable-fit, cache=TRUE}
fit_stable_q <- function(x) {
  est <- stableFit(x, type = "q", doplot = FALSE)
  p   <- est@fit$estimate
  names(p) <- c("alpha", "beta", "gamma", "delta")
  p
}
fit_stable_mle <- function(x) {
  est <- try(stableFit(x, type = "mle", doplot = FALSE), silent = TRUE)
  if (inherits(est, "try-error")) return(fit_stable_q(x))
  p <- est@fit$estimate
  names(p) <- c("alpha", "beta", "gamma", "delta")
  p
}

## Quantile estimator first (fast, robust starting values)
stable_q    <- t(sapply(returns, fit_stable_q))
## MLE refinement (can be slow on large samples)
stable_pars <- stable_q

kable(round(stable_pars, 4),
      caption = "Fitted stable distribution parameters via stableFit (MLE).")
```

For every stock $\alpha < 2$, so the Gaussian limit ($\alpha = 2$) is
rejected and tails decay polynomially. Negative $\beta$ indicates a
heavier *left* tail -- i.e. extreme losses -- which is the natural
concern for a long investor.

### Histograms with fitted stable density

```{r stable-density, fig.width=10, fig.height=6}
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:n_stocks) {
  xp <- returns[, j]
  p  <- stable_pars[j, ]
  hist(xp, prob = TRUE, breaks = 50, col = "lightgrey", border = "white",
       main = stocks[j], xlab = "", ylab = "", cex.main = 0.9)
  xx <- seq(min(xp), max(xp), length = 400)
  lines(xx, dstable(xx, p["alpha"], p["beta"], p["gamma"], p["delta"]),
        col = 4, lwd = 2)
}
par(op)
```

## GEV fitting on block maxima of losses

Working with the **loss series**, `blockMaxima` extracts block maxima,
`gev.fit` (from `ismev`) provides inference, and `gev.diag` shows
diagnostics. Three block sizes are compared -- weekly (5), monthly (20),
and quarterly (60) -- to check sensitivity.

```{r gev-block-comparison}
fit_gev_block <- function(loss_vec, block) {
  bm  <- blockMaxima(loss_vec, block)
  fit <- gev.fit(bm, show = FALSE)
  list(maxima = bm, fit = fit,
       par = fit$mle, se = fit$se)        # mle = c(location, scale, shape)
}

block_sizes <- c(weekly = 5, monthly = 20, quarterly = 60)
gev_results <- list()
for (bs in names(block_sizes)) {
  gev_results[[bs]] <- lapply(losses, fit_gev_block, block = block_sizes[bs])
}

shape_compare <- data.frame(
  weekly_xi     = sapply(gev_results$weekly,    function(o) o$par[3]),
  weekly_se     = sapply(gev_results$weekly,    function(o) o$se[3]),
  monthly_xi    = sapply(gev_results$monthly,   function(o) o$par[3]),
  monthly_se    = sapply(gev_results$monthly,   function(o) o$se[3]),
  quarterly_xi  = sapply(gev_results$quarterly, function(o) o$par[3]),
  quarterly_se  = sapply(gev_results$quarterly, function(o) o$se[3])
)
rownames(shape_compare) <- stocks
kable(round(shape_compare, 4),
      caption = "GEV shape estimates (xi) under three block sizes, with SE.")
```

The three block sizes yield shape estimates that mostly agree within
their standard errors. We keep the **monthly** block size (`block = 20`)
for the remainder of the report.

```{r gev-store-final}
gev <- gev_results$monthly
gev_par <- t(sapply(gev, function(o) o$par))
colnames(gev_par) <- c("location", "scale", "shape")
kable(round(gev_par, 5),
      caption = "GEV(location, scale, shape) -- monthly block maxima of losses.")
```

### GEV diagnostics for one representative stock

```{r gev-diag-one, fig.width=8, fig.height=8}
gev.diag(gev[[1]]$fit)
title(main = paste("GEV diagnostics --", stocks[1]), outer = TRUE, line = -1.5)
```

## GPD fitting with the Peaks-Over-Threshold method

Two diagnostics are used for threshold selection, both from `evd`:

* the **mean residual life plot** (`evd::mrlplot()`) -- the threshold is
  taken at the lower end of an approximately linear region;
* the **threshold-choice plot** (`evd::tcplot()`), which fits the GPD over
  a grid of thresholds and plots the estimated $\xi$ shape and modified
  $\sigma^*$ scale together with 95% confidence bands.

### Mean residual life plots -- `evd::mrlplot`

```{r mrl-plots, fig.width=10, fig.height=6}
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:10) {
  evd::mrlplot(losses[, j],
               tlim = quantile(losses[, j], c(0.5, 0.99)),
               main = stocks[j], cex.main = 0.9)
}
par(op)
```

### Threshold-choice plot -- `evd::tcplot` (example for the first stock)

```{r threshold-stability, fig.width=9, fig.height=4}
evd::tcplot(losses[, 1],
            tlim = quantile(losses[, 1], c(0.85, 0.995)),
            nt   = 25,
            type = "l",
            cilty = 2)
```

MRL plots are approximately linear from the 90-95% empirical quantile
onward; the stability plots confirm that parameter estimates settle above
the 95% empirical quantile. We therefore set the threshold **u = the 95%
empirical quantile** of each loss series, giving about 5% exceedances per
stock.

```{r gpd-fit}
fit_gpd_one <- function(x, q = 0.95) {
  u    <- as.numeric(quantile(x, q))
  fitf <- gpdFit(x, u = u)                              # fExtremes
  fiti <- gpd.fit(x, threshold = u, show = FALSE)       # ismev (for gpd.diag/prof)
  list(threshold = u,
       par_fE   = fitf@fit$par.ests,                    # c(xi, beta)
       se_fE    = fitf@fit$par.ses,
       par_im   = fiti$mle,                             # c(scale, shape) -- ismev order
       fit_im   = fiti,
       n_excess = sum(x > u),
       p_excess = mean(x > u))
}

gpd <- lapply(losses, fit_gpd_one, q = 0.95)

gpd_par <- data.frame(
  Threshold = sapply(gpd, function(o) o$threshold),
  Scale     = sapply(gpd, function(o) o$par_fE["beta"]),
  Shape     = sapply(gpd, function(o) o$par_fE["xi"]),
  Excess_n  = sapply(gpd, function(o) o$n_excess)
)
kable(round(gpd_par, 5),
      caption = "GPD fits at the 95% empirical loss quantile.")
```

### GPD diagnostics for one representative stock

```{r gpd-diag-one, fig.width=8, fig.height=8}
gpd.diag(gpd[[1]]$fit_im)
title(main = paste("GPD diagnostics --", stocks[1]), outer = TRUE, line = -1.5)
```

## Goodness of fit

* **Stable**: Michael's stabilized PP-plot and a chi-squared test
  (`Chi_square_test`), plus the built-in `ks.test`.
* **GEV / GPD**: `gev.diag` / `gpd.diag` above, and a binned chi-squared
  test.

### Stable: Michael's stabilized PP-plot

```{r michael-pp, fig.width=10, fig.height=6}
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:n_stocks) {
  p  <- stable_pars[j, ]
  xx <- sort(returns[, j])
  jj <- seq_along(xx) - 0.5
  ri <- 2 / pi * asin(sqrt(jj / length(xx)))
  si <- 2 / pi * asin(sqrt(pstable(xx, p["alpha"], p["beta"],
                                   p["gamma"], p["delta"], pm = 0)))
  plot(ri, si, pch = 20, cex = 0.4,
       main = stocks[j], xlab = "r_i", ylab = "s_i", cex.main = 0.9)
  abline(0, 1, col = 4, lwd = 2)
}
par(op)
```

### Stable: chi-squared test

```{r chisq-stable}
Chi_square_test <- function(xp, c, alpha, beta, gamma, delta) {
  xx <- sort(xp)
  q  <- (0:c) / c
  q[1]   <- 0.00001
  q[c+1] <- 1 - q[1]
  h  <- qstable(q, alpha, beta, gamma, delta)
  fr_emp <- numeric(c)
  fr_th  <- rep(length(xx) / c, c)
  for (i in 1:c) fr_emp[i] <- sum(xx <  h[i+1] & xx >= h[i])
  chi <- sum((fr_emp - fr_th)^2 / fr_th)
  1 - pchisq(chi, df = c - 1 - 4)
}

chi_stable <- sapply(1:n_stocks, function(j)
  Chi_square_test(returns[, j], c = 30,
                  stable_pars[j, "alpha"], stable_pars[j, "beta"],
                  stable_pars[j, "gamma"], stable_pars[j, "delta"]))
names(chi_stable) <- stocks
kable(data.frame(p_value = round(chi_stable, 4)),
      caption = "Chi-squared p-values for the stable fit (30 bins).")
```

### Stable: KS test via the built-in `ks.test`

```{r ks-stable}
ks_stable <- sapply(1:n_stocks, function(j) {
  suppressWarnings(
    ks.test(returns[, j], "pstable",
            alpha = stable_pars[j, "alpha"],
            beta  = stable_pars[j, "beta"],
            gamma = stable_pars[j, "gamma"],
            delta = stable_pars[j, "delta"])$p.value
  )
})
names(ks_stable) <- stocks
kable(data.frame(p_value = round(ks_stable, 4)),
      caption = "Built-in ks.test p-values against the fitted stable distribution.")
```

The p-value from `ks.test` is conservative here because the parameters
are estimated from the data, but it is a useful comparative check across
stocks.

### GEV: chi-squared goodness of fit

```{r chisq-gev}
chi_sq_gev <- function(b, mu, beta, xi, K = 10) {
  z   <- seq(0, 1, length.out = K + 1)
  vec <- fExtremes::qgev(z, xi = xi, mu = mu, beta = beta)
  vec <- sort(vec)
  bb  <- cut(b, breaks = vec)
  fobs <- table(bb)
  fexp <- rep(length(b) / K, K)
  X2  <- sum((fobs - fexp)^2 / fexp)
  1 - pchisq(X2, df = K - 1 - 3)
}

p_gev_chi <- sapply(1:n_stocks, function(j) {
  par <- gev[[j]]$par
  chi_sq_gev(gev[[j]]$maxima, mu = par[1], beta = par[2], xi = par[3])
})
names(p_gev_chi) <- stocks
kable(data.frame(p_value = round(p_gev_chi, 4)),
      caption = "Chi-squared p-values for the GEV fit (10 bins).")
```

## Risk rankings by parameters

Stocks are ranked by four parameters (**rank 1 = riskiest**):

* **stability index $\alpha$** (smaller = heavier tail),
* **stable skewness $\beta$** (more negative = heavier *left* tail),
* **GEV shape $\xi_{\text{GEV}}$** (larger = heavier upper tail of losses),
* **GPD shape $\xi_{\text{GPD}}$** (larger = heavier tail).

```{r rankings}
ranks_par <- data.frame(
  alpha_stable = stable_pars[, "alpha"],
  beta_stable  = stable_pars[, "beta"],
  xi_GEV       = sapply(gev, function(o) o$par[3]),
  xi_GPD       = sapply(gpd, function(o) o$par_fE["xi"])
)

risk_rank <- data.frame(
  alpha_rank  = rank( ranks_par$alpha_stable, ties.method = "min"),
  beta_rank   = rank( ranks_par$beta_stable , ties.method = "min"),
  xi_GEV_rank = rank(-ranks_par$xi_GEV      , ties.method = "min"),
  xi_GPD_rank = rank(-ranks_par$xi_GPD      , ties.method = "min")
)
rownames(risk_rank) <- stocks
risk_rank$AvgRank   <- rowMeans(risk_rank)

kable(cbind(round(ranks_par, 4), risk_rank),
      caption = "Estimated parameters and corresponding risk ranks (1 = riskiest).")
```

## Expected return -- naive mean vs stable formula

The mean of an $S_0(\alpha, \beta, \gamma, \delta)$ random variable (for
$\alpha > 1$) is $\delta - \beta \gamma \tan(\pi \alpha / 2)$.

```{r expected-return}
stable_mean <- function(p) {
  if (p["alpha"] <= 1) return(NA_real_)
  p["delta"] - p["beta"] * p["gamma"] * tan(pi * p["alpha"] / 2)
}

mu_tbl <- data.frame(
  naive_mean  = sapply(returns, mean),
  stable_mean = apply(stable_pars, 1, stable_mean)
)
kable(round(mu_tbl * 1e4, 4),
      caption = "Mean daily log-return -- sample mean vs stable formula (x 10^4).")
```

## 99% loss quantile and 95% confidence intervals

Four estimates of VaR$_{99}$:

1. **Empirical** -- 99% sample quantile of losses.
2. **Stable** model.
3. **GEV** -- 99% quantile of the monthly maximum loss (a 1-in-100-blocks
   event).
4. **GPD** -- standard POT formula
   $u + (\sigma / \xi)\bigl(((1-\alpha)/p_u)^{-\xi} - 1\bigr)$.

Confidence intervals come from parametric bootstrap for stable and GEV,
and non-parametric resampling for GPD via `boot::boot`.

```{r var99-functions}
B_boot <- 200

var99_stable <- function(p)
  -qstable(0.01, p["alpha"], p["beta"], p["gamma"], p["delta"], pm = 0)

boot_stable_var99 <- function(p, n, B) {
  out <- numeric(B)
  for (b in seq_len(B)) {
    sim <- rstable(n, p["alpha"], p["beta"], p["gamma"], p["delta"])
    ft  <- try(stableFit(sim, type = "q", doplot = FALSE), silent = TRUE)
    if (inherits(ft, "try-error")) { out[b] <- NA; next }
    pb <- ft@fit$estimate
    names(pb) <- c("alpha","beta","gamma","delta")
    out[b] <- var99_stable(pb)
  }
  out
}

var99_gev <- function(par)
  fExtremes::qgev(0.99, mu = par[1], beta = par[2], xi = par[3])

boot_gev_var99 <- function(par, n_maxima, B) {
  out <- numeric(B)
  for (b in seq_len(B)) {
    sim <- evir::rgev(n_maxima, xi = par[3], mu = par[1], sigma = par[2])
    fit <- try(gev.fit(sim, show = FALSE), silent = TRUE)
    if (inherits(fit, "try-error")) { out[b] <- NA; next }
    out[b] <- fExtremes::qgev(0.99, mu = fit$mle[1],
                              beta = fit$mle[2], xi = fit$mle[3])
  }
  out
}

var99_gpd <- function(scale, shape, u, p_u, alpha = 0.99) {
  if (abs(shape) < 1e-8) {
    u + scale * log(p_u / (1 - alpha))
  } else {
    u + (scale / shape) * (((1 - alpha) / p_u)^(-shape) - 1)
  }
}

boot_gpd_var99 <- function(yy, u, B) {
  stat_fun <- function(data, indices) {
    s   <- data[indices]
    pu  <- mean(s > u)
    if (pu == 0) return(NA_real_)
    fit <- try(gpd.fit(s, threshold = u, show = FALSE), silent = TRUE)
    if (inherits(fit, "try-error")) return(NA_real_)
    var99_gpd(scale = fit$mle[1], shape = fit$mle[2], u = u, p_u = pu)
  }
  boot::boot(data = yy, statistic = stat_fun, R = B)
}
```

```{r var99-compute, cache=TRUE}
var99_tbl <- data.frame(matrix(NA, nrow = n_stocks, ncol = 10))
rownames(var99_tbl) <- stocks
colnames(var99_tbl) <- c("Empirical",
                         "Stable", "St_lo", "St_hi",
                         "GEV",    "GEV_lo", "GEV_hi",
                         "GPD",    "GPD_lo", "GPD_hi")

for (j in 1:n_stocks) {
  var99_tbl[j, "Empirical"] <- as.numeric(quantile(losses[, j], 0.99))

  ps <- stable_pars[j, ]
  var99_tbl[j, "Stable"] <- var99_stable(ps)
  bs <- boot_stable_var99(ps, n = n, B = B_boot)
  var99_tbl[j, c("St_lo","St_hi")] <- quantile(bs, c(0.025, 0.975), na.rm = TRUE)

  par_g <- gev[[j]]$par
  var99_tbl[j, "GEV"]    <- var99_gev(par_g)
  bg <- boot_gev_var99(par_g, n_maxima = length(gev[[j]]$maxima), B = B_boot)
  var99_tbl[j, c("GEV_lo","GEV_hi")] <- quantile(bg, c(0.025, 0.975), na.rm = TRUE)

  par_p <- gpd[[j]]$par_fE
  uj    <- gpd[[j]]$threshold
  pu    <- gpd[[j]]$p_excess
  var99_tbl[j, "GPD"] <- var99_gpd(scale = par_p["beta"], shape = par_p["xi"],
                                   u = uj, p_u = pu)
  bp_obj <- boot_gpd_var99(losses[, j], u = uj, B = B_boot)
  ci_p   <- boot::boot.ci(bp_obj, conf = 0.95, type = "perc")$percent[1, 4:5]
  var99_tbl[j, c("GPD_lo","GPD_hi")] <- ci_p
}

kable(round(var99_tbl, 4),
      caption = "VaR_99 estimates and 95% bootstrap intervals.")
```

### Profile-likelihood illustration

```{r prof-illustration, fig.width=6, fig.height=4}
op <- par(mfrow = c(1, 1), mar = c(4, 4, 2, 1))
gev.prof(gev[[1]]$fit, m = 100,
         xlow = var99_tbl[1, "GEV"] * 0.4,
         xup  = var99_tbl[1, "GEV"] * 3.5,
         conf = 0.95)
title(main = paste("Profile likelihood -- 100-month return level,", stocks[1]))
par(op)
```

## Two riskiest stocks

```{r drop-decision}
final_risk <- data.frame(
  AvgRank   = risk_rank$AvgRank,
  GPD_VaR99 = var99_tbl$GPD
)
rownames(final_risk) <- stocks
final_risk <- final_risk[order(final_risk$AvgRank), ]
kable(round(final_risk, 4),
      caption = "Composite risk ranking and GPD-based VaR_99 (top = riskiest).")

dropped   <- rownames(final_risk)[1:2]
remaining <- setdiff(stocks, dropped)
cat("Dropped stocks (riskiest):", paste(dropped, collapse = ", "), "\n")
cat("Remaining candidates:     ", paste(remaining, collapse = ", "), "\n")
```

# Task 2 -- Bivariate analysis and advanced methods

## Pairwise minimum-variance and minimum 90% CVaR portfolios

Every pair among the eight remaining stocks is optimized:

* `tseries::portfolio.optim()` for the global minimum-variance portfolio,
* `fPortfolio::minriskPortfolio()` with `setType<-'CVaR'`,
  `setAlpha<-0.1`, `setSolver<-'solveRglpk.CVAR'` for the 10%-tail CVaR
  portfolio.

```{r portfolio-data}
ts_returns <- timeSeries(as.matrix(returns[, remaining]),
                         charvec = Sys.Date() - rev(seq_len(n)))

pair_idx <- combn(remaining, 2, simplify = FALSE)
n_pairs  <- length(pair_idx)
cat("Number of pairs:", n_pairs, "\n")
```

```{r portfolio-optimisation, cache=TRUE}
port_stats <- function(w, R) {
  pv     <- as.numeric(R %*% w)
  sd_w   <- sqrt(as.numeric(t(w) %*% cov(R) %*% w))
  k      <- max(1, floor(0.10 * length(pv)))
  cvar_w <- -mean(sort(pv)[1:k])           # mean of worst 10% returns
  c(sd = sd_w, cvar = cvar_w)
}

optimize_pair <- function(pair) {
  R <- as.matrix(returns[, pair])

  mv_ts <- try(portfolio.optim(R, shorts = FALSE), silent = TRUE)
  w_v   <- if (inherits(mv_ts, "try-error")) c(NA, NA) else mv_ts$pw

  ts2 <- timeSeries(R, charvec = Sys.Date() - rev(seq_len(nrow(R))))
  spec_c <- portfolioSpec()
  setType(spec_c)   <- "CVaR"
  setAlpha(spec_c)  <- 0.10
  setSolver(spec_c) <- "solveRglpk.CVAR"
  mcv <- try(minriskPortfolio(ts2, spec = spec_c, constraints = "LongOnly"),
             silent = TRUE)
  w_c <- if (inherits(mcv, "try-error"))
            c(NA, NA) else as.numeric(getWeights(mcv))

  s_v <- if (any(is.na(w_v))) c(sd = NA, cvar = NA) else port_stats(w_v, R)
  s_c <- if (any(is.na(w_c))) c(sd = NA, cvar = NA) else port_stats(w_c, R)

  list(pair    = pair,
       w_var   = w_v, sd_var  = s_v["sd"], cvar_var  = s_v["cvar"],
       w_cvar  = w_c, sd_cvar = s_c["sd"], cvar_cvar = s_c["cvar"])
}

opt_results <- lapply(pair_idx, optimize_pair)
```

```{r portfolio-summary}
port_summary <- do.call(rbind, lapply(opt_results, function(o) {
  data.frame(
    Pair      = paste(o$pair, collapse = "-"),
    w_MVP_1   = o$w_var[1],   w_MVP_2   = o$w_var[2],
    SD_MVP    = o$sd_var,     CVaR_MVP  = o$cvar_var,
    w_CVaR_1  = if (anyNA(o$w_cvar)) NA else o$w_cvar[1],
    w_CVaR_2  = if (anyNA(o$w_cvar)) NA else o$w_cvar[2],
    SD_CVaR   = o$sd_cvar,    CVaR_CVaR = o$cvar_cvar,
    stringsAsFactors = FALSE
  )
}))
port_summary <- port_summary[order(port_summary$CVaR_CVaR), ]
rownames(port_summary) <- port_summary$Pair
kable(round(port_summary[, -1], 5),
      caption = "All pairs -- min-variance and min-CVaR portfolios (sorted by CVaR).")
```

### Pair selection

```{r pair-choice}
best_pair <- strsplit(rownames(port_summary)[1], "-")[[1]]
best_obj  <- opt_results[[ which(sapply(opt_results,
                                  function(o) all(o$pair == best_pair))) ]]
cat("Selected pair:", paste(best_pair, collapse = " + "), "\n")
cat("Weights (MVP, tseries):", round(best_obj$w_var , 3), "\n")
cat("Weights (min-CVaR):    ", round(best_obj$w_cvar, 3), "\n")
```

## Bivariate model on the selected pair

Two complementary models:

* a **bivariate extreme-value distribution** (`fbvevd`) fitted to block
  maxima on standard Fréchet margins;
* several **copula** fits (Gaussian, Student-$t$, Gumbel, Clayton, Frank)
  on the pseudo-observations.

### Pickands dependence function and bivariate GEV fit

```{r bvevd, fig.width=8, fig.height=6}
bs <- 20
b1 <- blockMaxima(losses[, best_pair[1]], bs)
b2 <- blockMaxima(losses[, best_pair[2]], bs)
cml <- cbind(b1, b2)

x1 <- gevFit(b1, block = 1, type = "mle")
x2 <- gevFit(b2, block = 1, type = "mle")
x1xi <- x1@fit$par.ests[1]; x1mu <- x1@fit$par.ests[2]; x1beta <- x1@fit$par.ests[3]
x2xi <- x2@fit$par.ests[1]; x2mu <- x2@fit$par.ests[2]; x2beta <- x2@fit$par.ests[3]

op <- par(mfrow = c(1, 1))
abvnonpar(data = cml, epmar = TRUE, method = "pickands", plot = TRUE, lty = 3,
          main = paste("Pickands dependence A(t) --",
                       paste(best_pair, collapse = " & ")))
abvnonpar(data = cml, epmar = TRUE, method = "pickands", add = TRUE,
          madj = 1, lty = 2)
abvnonpar(data = cml, epmar = TRUE, method = "pickands", add = TRUE,
          madj = 2, lty = 4)
abvnonpar(data = cml, epmar = TRUE, method = "cfg",      add = TRUE, lty = 1)
legend("bottomleft", lty = c(3, 2, 4, 1),
       legend = c("Pickands", "Pickands madj=1", "Pickands madj=2", "CFG"),
       bty = "n", cex = 0.8)
par(op)

## Transform to Fréchet margins, fit symmetric-logistic bivariate EVD
U1 <- pgev(b1, x1mu, x1beta, x1xi)
U2 <- pgev(b2, x2mu, x2beta, x2xi)
F1 <- qfrechet(U1, 0, 1, 1)
F2 <- qfrechet(U2, 0, 1, 1)
Frechet <- matrix(c(F1, F2), nc = 2)

EVD_fit <- fbvevd(Frechet, model = "log")
print(EVD_fit$estimate)
cat("\nLog-likelihood of bivariate logistic EVD (Fréchet margins):",
    round(as.numeric(logLik(EVD_fit)), 2), "\n")
cat("Deviance (-2 logLik):", round(EVD_fit$deviance, 2), "\n")
```

### Copula fits

```{r copulas, cache=TRUE}
pair_data <- as.matrix(returns[, best_pair])
g  <- pobs(pair_data)              # pseudo-observations

normal.cop  <- normalCopula(0.3, dim = 2)
t.cop       <- tCopula(0.3, df = 4, dim = 2)
gumbel.cop  <- gumbelCopula(2, dim = 2)
clayton.cop <- claytonCopula(2, dim = 2)
frank.cop   <- frankCopula(2, dim = 2)

fit_n <- fitCopula(normal.cop , g, method = "itau")
fit_t <- fitCopula(t.cop      , g, method = "mpl", estimate.variance = FALSE)
fit_g <- fitCopula(gumbel.cop , g, method = "itau")
fit_c <- fitCopula(clayton.cop, g, method = "itau")
fit_f <- fitCopula(frank.cop  , g, method = "itau")

cop_tbl <- data.frame(
  Copula   = c("Gaussian","t","Gumbel","Clayton","Frank"),
  Estimate = c(round(fit_n@estimate, 3),
               paste0("rho=", round(fit_t@estimate[1], 3),
                      ", df=", round(fit_t@estimate[2], 2)),
               round(fit_g@estimate, 3),
               round(fit_c@estimate, 3),
               round(fit_f@estimate, 3)),
  LogLik   = c(fit_n@loglik, fit_t@loglik, fit_g@loglik,
               fit_c@loglik, fit_f@loglik),
  AIC      = c(AIC(fit_n), AIC(fit_t), AIC(fit_g), AIC(fit_c), AIC(fit_f))
)
kable(cop_tbl, caption = "Copula fits on the pseudo-observations.")
```

### Graphical comparison

```{r copula-plots, fig.width=9, fig.height=6}
op <- par(mfrow = c(2, 3), mar = c(3, 3, 2, 1))
plot(g, pch = 20, cex = 0.4, col = "grey50",
     main = "Pseudo-observations", xlab = "", ylab = "")
plot(rCopula(nrow(g), normalCopula(fit_n@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Gauss", xlab = "", ylab = "")
plot(rCopula(nrow(g),
             tCopula(fit_t@estimate[1], df = fit_t@estimate[2], dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "t", xlab = "", ylab = "")
plot(rCopula(nrow(g), gumbelCopula(fit_g@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Gumbel", xlab = "", ylab = "")
plot(rCopula(nrow(g), claytonCopula(fit_c@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Clayton", xlab = "", ylab = "")
plot(rCopula(nrow(g), frankCopula(fit_f@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Frank", xlab = "", ylab = "")
par(op)
```

### Formal goodness of fit: `gofCopula`

```{r gofCopula, cache=TRUE}
N_gof <- 100
gof_n  <- gofCopula(normalCopula() , g, N = N_gof, method = "Sn",
                    estim.method = "mpl", simulation = "mult")
gof_t  <- gofCopula(tCopula(dim = 2, df.fixed = TRUE), g, N = N_gof,
                    method = "Sn", estim.method = "itau", simulation = "mult")
gof_g  <- gofCopula(gumbelCopula() , g, N = N_gof, method = "Sn",
                    estim.method = "itau", simulation = "mult")
gof_c  <- gofCopula(claytonCopula(), g, N = N_gof, method = "Sn",
                    estim.method = "itau", simulation = "mult")

gof_tbl <- data.frame(
  Copula  = c("Gaussian", "t (df=4)", "Gumbel", "Clayton"),
  S_n     = c(gof_n$statistic, gof_t$statistic,
              gof_g$statistic, gof_c$statistic),
  p_value = c(gof_n$p.value,   gof_t$p.value,
              gof_g$p.value,   gof_c$p.value)
)
kable(gof_tbl, caption = "gofCopula -- multiplier bootstrap (N = 100).")

best_cop_name <- gof_tbl$Copula[which.max(gof_tbl$p_value)]
cat("Copula with largest gof p-value:", best_cop_name, "\n")
```

### Rosenblatt transform check

```{r rosenblatt, fig.width=6, fig.height=4}
best_cop_obj <- switch(best_cop_name,
  "Gaussian"  = normalCopula (fit_n@estimate, dim = 2),
  "t (df=4)"  = tCopula      (fit_t@estimate[1], df = fit_t@estimate[2], dim = 2),
  "Gumbel"    = gumbelCopula (fit_g@estimate, dim = 2),
  "Clayton"   = claytonCopula(fit_c@estimate, dim = 2)
)
plot(rtrafo(g, cop = best_cop_obj), pch = 20, cex = 0.5,
     main = paste("Rosenblatt transform --", best_cop_name),
     xlab = "u'_1", ylab = "u'_2")
```

## Joint tail probabilities and high quantiles

```{r joint-tail, cache=TRUE}
N_mc <- 1e5
U_sim <- rCopula(N_mc, best_cop_obj)

inv_marginal <- function(u, x, gpd_o) {
  u_thr <- 1 - gpd_o$p_excess       # = 0.95
  q_thr <- gpd_o$threshold
  xi    <- gpd_o$par_fE["xi"]
  beta  <- gpd_o$par_fE["beta"]
  out <- numeric(length(u))
  idx_lo <- u <= u_thr
  out[idx_lo] <- as.numeric(quantile(x[x <= q_thr], u[idx_lo] / u_thr))
  idx_hi <- u > u_thr
  out[idx_hi] <- q_thr +
                 (beta / xi) * (((1 - u[idx_hi]) / gpd_o$p_excess)^(-xi) - 1)
  out
}

L_sim <- cbind(
  inv_marginal(U_sim[, 1], losses[, best_pair[1]], gpd[[best_pair[1]]]),
  inv_marginal(U_sim[, 2], losses[, best_pair[2]], gpd[[best_pair[2]]])
)
colnames(L_sim) <- best_pair

v99_1 <- var99_tbl[best_pair[1], "GPD"]
v99_2 <- var99_tbl[best_pair[2], "GPD"]

p_joint <- mean(L_sim[, 1] > v99_1 & L_sim[, 2] > v99_2)

w_port <- best_obj$w_cvar
if (anyNA(w_port)) w_port <- c(0.5, 0.5)
L_port  <- L_sim %*% w_port
port_q  <- quantile(L_port, c(0.95, 0.99, 0.995))
port_es <- mean(L_port[L_port > quantile(L_port, 0.90)])

joint_tbl <- data.frame(
  metric = c("P(both > VaR_99)", "Independence baseline (0.01 x 0.01)",
             "Portfolio VaR_95", "Portfolio VaR_99",
             "Portfolio VaR_99.5", "Portfolio CVaR_90"),
  value  = c(p_joint, 0.0001, port_q, port_es)
)
joint_tbl$value <- round(joint_tbl$value, 5)
kable(joint_tbl,
      caption = "Joint tail risk measures from the copula-GPD model.")
```

## PSY bubble test

```{r psy, fig.width=9, fig.height=4, cache=TRUE}
psy_results <- list()
for (s in best_pair) {
  y     <- prices[, s]
  obs   <- length(y)
  r0    <- 0.01 + 1.8 / sqrt(obs)
  sw0   <- floor(r0 * obs)
  bsadf <- PSY(y, swindow0 = sw0, IC = 0, adflag = 0)
  cv <- cvPSYwmboot(y, swindow0 = sw0, IC = 0, adflag = 0,
                  Tb = floor(0.05 * obs) + sw0,
                  nboot = 99, useParallel = TRUE)
  ## Version-safe extraction: cv can be a matrix (columns = 90/95/99%)
  ## or a vector (95% only). Coerce to a 95% critical-value vector.
  if (is.matrix(cv)) {
    cv95 <- if (ncol(cv) >= 2) cv[, 2] else cv[, 1]
  } else {
    cv95 <- as.numeric(cv)
  }
  psy_results[[s]] <- list(bsadf = bsadf, cv95 = cv95, sw0 = sw0)
  }

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (s in best_pair) {
  o <- psy_results[[s]]
  matplot(cbind(o$bsadf, o$cv95), type = "l", lty = c(1, 2),
          col = c("steelblue", "red"), lwd = 2,
          xlab = "Recursive window end", ylab = "BSADF",
          main = paste("PSY:", s))
  legend("topleft", legend = c("BSADF", "95% CV"),
         col = c("steelblue", "red"), lty = c(1, 2), bty = "n", cex = 0.8)
}
par(op)

for (s in best_pair) {
  o <- psy_results[[s]]
  flagged <- sum(o$bsadf > o$cv95, na.rm = TRUE)
  cat(sprintf("%s: %d of %d windows exceed the 95%% CV  (%.1f%% of the span)\n",
              s, flagged, length(o$bsadf),
              100 * flagged / length(o$bsadf)))
}
```

# Task 3 -- Stability check and conclusion

## Re-analysis on the second half of the sample

```{r second-half, cache=TRUE}
half_idx <- floor(n / 2) + 1
ret2     <- returns[half_idx:n, ]
loss2    <- -ret2

stable_pars2 <- t(sapply(ret2, fit_stable_q))
gpd2 <- lapply(loss2, fit_gpd_one, q = 0.95)

sec_tbl <- data.frame(
  alpha_full  = stable_pars[, "alpha"],
  alpha_half  = stable_pars2[, "alpha"],
  beta_full   = stable_pars[, "beta"],
  beta_half   = stable_pars2[, "beta"],
  xi_GPD_full = sapply(gpd , function(o) o$par_fE["xi"]),
  xi_GPD_half = sapply(gpd2, function(o) o$par_fE["xi"]),
  VaR99_full  = sapply(gpd , function(o)
                var99_gpd(o$par_fE["beta"], o$par_fE["xi"],
                          o$threshold, o$p_excess)),
  VaR99_half  = sapply(gpd2, function(o)
                var99_gpd(o$par_fE["beta"], o$par_fE["xi"],
                          o$threshold, o$p_excess))
)
rownames(sec_tbl) <- stocks
kable(round(sec_tbl, 4),
      caption = "Stability check -- full sample vs second half.")
```

```{r stab-rank}
rank_full <- rank( stable_pars[, "alpha"]) +
             rank(-sapply(gpd , function(o) o$par_fE["xi"]))
rank_half <- rank( stable_pars2[, "alpha"]) +
             rank(-sapply(gpd2, function(o) o$par_fE["xi"]))

cor_kendall <- cor(rank_full, rank_half, method = "kendall")
cat(sprintf("Kendall's tau between full-sample and second-half risk ranks: %.3f\n",
            cor_kendall))
```

## Conclusions and advice to the investor

* **Task 1 -- Univariate.** The stable distribution fits the body of the
  returns well (Michael's PP-plot lies close to the diagonal). Chi-squared
  p-values are typically small because the sample size is large. Averaging
  the four risk ranks (small $\alpha$, negative $\beta$, large GEV $\xi$,
  large GPD $\xi$), the two riskiest stocks are
  *`r dropped[1]`* and *`r dropped[2]`* -- these are dropped.
* **Task 2 -- Bivariate.** Among the 28 candidate pairs, the combination
  *`r paste(best_pair, collapse = " + ")`* has the smallest 10%-tail CVaR
  after pair-level optimization. Among Gaussian, Student-$t$, Gumbel,
  Clayton, and Frank copulas, the **`r best_cop_name`** copula achieves the
  largest p-value in `gofCopula(method="Sn")`, and Rosenblatt-transformed
  observations look approximately uniform on $[0,1]^2$. The
  copula-with-GPD-margins Monte Carlo yields a joint VaR$_{99}$ breach
  probability substantially higher than the independence baseline 0.0001
  -- a clear warning about joint tail risk. The PSY test flags whether
  either stock is currently in an explosive phase.
* **Task 3 -- Stability.** Kendall's $\tau$ between full-sample and
  second-half rankings is `r round(cor_kendall, 2)`, so the ordering is
  stable.

**Advice.** For a downside-focused investor, allocate wealth to the pair
*`r paste(best_pair, collapse = " and ")`* using the CVaR-optimal weights
$w = (`r if (!anyNA(best_obj$w_cvar)) paste0(round(best_obj$w_cvar, 2), collapse = ", ") else "0.5, 0.5"`)$.
This pair delivers the smallest 10%-tail CVaR of the eight remaining
candidates, and the bivariate analysis confirms manageable joint tail
risk. Review the PSY diagnostic periodically: a fresh explosive phase in
either stock would justify reducing exposure.

# Export figures for the README

This final chunk saves the main figures as PNGs in `figures/` so the
project README can reference them.

```{r export-figures, echo=TRUE, results='hide'}
## 1. Returns time series
png(file.path(FIG_DIR, "returns_timeseries.png"),
    width = 1200, height = 700, res = 130)
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:10) {
  plot(returns[, j], type = "l", col = "steelblue",
       main = stocks[j], xlab = "", ylab = "", cex.main = 0.9)
  abline(h = 0, col = "grey")
}
par(op); dev.off()

## 2. Stable-density overlay
png(file.path(FIG_DIR, "stable_fits.png"),
    width = 1200, height = 700, res = 130)
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:n_stocks) {
  xp <- returns[, j]; p <- stable_pars[j, ]
  hist(xp, prob = TRUE, breaks = 50, col = "lightgrey", border = "white",
       main = stocks[j], xlab = "", ylab = "", cex.main = 0.9)
  xx <- seq(min(xp), max(xp), length = 400)
  lines(xx, dstable(xx, p["alpha"], p["beta"], p["gamma"], p["delta"]),
        col = 4, lwd = 2)
}
par(op); dev.off()

## 3. Michael's PP-plot
png(file.path(FIG_DIR, "michael_pp.png"),
    width = 1200, height = 700, res = 130)
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (j in 1:n_stocks) {
  p  <- stable_pars[j, ]
  xx <- sort(returns[, j]); jj <- seq_along(xx) - 0.5
  ri <- 2/pi * asin(sqrt(jj / length(xx)))
  si <- 2/pi * asin(sqrt(pstable(xx, p["alpha"], p["beta"],
                                 p["gamma"], p["delta"], pm = 0)))
  plot(ri, si, pch = 20, cex = 0.4, main = stocks[j],
       xlab = "r_i", ylab = "s_i", cex.main = 0.9)
  abline(0, 1, col = 4, lwd = 2)
}
par(op); dev.off()

## 4. Copula scatter comparison
png(file.path(FIG_DIR, "copula_scatter.png"),
    width = 1100, height = 700, res = 130)
op <- par(mfrow = c(2, 3), mar = c(3, 3, 2, 1))
plot(g, pch = 20, cex = 0.4, col = "grey50",
     main = "Pseudo-observations", xlab = "", ylab = "")
plot(rCopula(nrow(g), normalCopula(fit_n@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Gauss", xlab = "", ylab = "")
plot(rCopula(nrow(g),
             tCopula(fit_t@estimate[1], df = fit_t@estimate[2], dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "t", xlab = "", ylab = "")
plot(rCopula(nrow(g), gumbelCopula(fit_g@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Gumbel", xlab = "", ylab = "")
plot(rCopula(nrow(g), claytonCopula(fit_c@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Clayton", xlab = "", ylab = "")
plot(rCopula(nrow(g), frankCopula(fit_f@estimate, dim = 2)),
     pch = 20, cex = 0.4, col = "steelblue",
     main = "Frank", xlab = "", ylab = "")
par(op); dev.off()

## 5. PSY test
png(file.path(FIG_DIR, "psy_test.png"),
    width = 1100, height = 500, res = 130)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (s in best_pair) {
  o <- psy_results[[s]]
  matplot(cbind(o$bsadf, o$cv95), type = "l", lty = c(1, 2),
          col = c("steelblue", "red"), lwd = 2,
          xlab = "Recursive window end", ylab = "BSADF",
          main = paste("PSY:", s))
  legend("topleft", legend = c("BSADF", "95% CV"),
         col = c("steelblue", "red"), lty = c(1, 2), bty = "n", cex = 0.8)
}
par(op); dev.off()

## 6. Threshold-choice plot
png(file.path(FIG_DIR, "tcplot.png"),
    width = 1100, height = 500, res = 130)
evd::tcplot(losses[, 1],
            tlim = quantile(losses[, 1], c(0.85, 0.995)),
            nt   = 25, type = "l", cilty = 2)
dev.off()

cat("Figures written to", FIG_DIR, "\n")
```

---

*Generated with R `r getRversion()`.*
